Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-04 intro.tex
edits to intro,related and methods

\section{Introduction}\label{sec:intro}

@@

Complicated time series are ubiquitous in modern scientific research. 
%Characterizing and forecasting the future behavior of these observed systems is a vastly researched area with many open questions. 
Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}\footnote{An approx of (blah ) entropy which we make more rigorous in blah}. On the low end of that spectrum are time series that exhibit perfect predictive structure, i.e, signals whose future value are perfectly predicted by knowledge of past values from some ``ideal" model. In particular, there is a an underlying process that generates and transmits information from the past to the future in a perfectly predictable fashion. Some examples of this are constant or low-period signals. On the opposite end of this spectrum are signals which are ``fully complex", i.e., the underlying generating process creates information so rapidly that no information at all is transmitted from the past to the future. Some examples, of this are white noise or random walk processes. With signals on this side of the spectrum, knowledge of the past gives absolutely no insight into the future, \emph{regardless} of the chosen model.  In the middle of this spectrum are where complicated \emph{and} interesting---from a modeling perspective---signals exist, e.g., observations of a deterministic chaotic trajectory, high period signals with some level of noise on top, etc. With time series in this portion of the spectrum, enough information is being transmitted from the past to the future that given the \emph{ideal} model the future of the observed system could be forecast with high accuracy.

Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{weigend-book}, let alone an arbitrary time series. This naturally leads to an important and hard question: given a complicated real-valued time series, is it possible to reliably quantify where on this spectrum the time series lies? i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast using a prediction strategy based on predictive structure (temporal forward information transfer)? We define this spectrum as well as what is meant by low, medium and high rigorously in Section \ref{sec:meaComplex}.

To answer this question and for the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which appear \emph{complicated}, i.e., they exhibit interesting behavior, for instance, not of low period, strictly monotonic, or constant. We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
We argue that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy of a real-valued-finite-length-noisy time series through ordinal analysis---is an effective way to explore that conjecture. In light of this, for the purposes of this paper we define \emph{complexity} as an approximation of Shannon \emph{entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}. In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity.

@@

To explore this conjecture we require a test bed for generating a broad array of time series to analyze which covers the full spectrum of complexity. We argue that the ideal test bed for this purpose are computer performances, e.g., the processor efficiency during the execution of a particular program. Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor efficiency highly complicated. 
%Accurate forecasts of these quantities, if one could
%construct them, could be used to improve computer design. 
% If one
%could predict that a particular computational thread %would be bogged
%down for the next 0.6 seconds waiting for data from main memory, for
%instance, one could save power by putting that thread %on hold for that
%time period (e.g., by migrating it to a processing %unit whose clock
%speed is scaled back).

As a result computer performance traces are very
complicated and range across the entire spectrum of complexity, making this an ideal test bed.  Even a simple ``microkernel," like a three-line loop that
repeatedly initializes the upper triangle of a matrix in column-major order, can produce everything from periodic to
{\sl chaotic} performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
Figure~\ref{fig:ipc}.%, and chaos places fundamental limits on predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of \col, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
More interesting programs, e.g., compilers and numerical solvers, produce time series which exist throughout the spectrum of complexity: from completely predictable to completely unstructured. In addition, some performance traces even exhibit interesting regime changes as a program switches between varying subroutines. With a single system (computer performance) capable of producing complicated time series that vary so greatly in complexity (periodic to noise), it would be invaluable to known a priori if a trace contained enough predictive structure to build a  forecast model based on structure before time and effort was spent on trying to model and forecast it.

@@

This paper provides insight into when, why, and how forecast strategies fail when they are applied to complicated time series by empirically estimating their inherent complexity. We focus here on the specific example of
computer performance but believe the results apply to a much broader class of time series. We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity, as well as from
measurement issues like noise, aggregation, and finite data
length---is \emph{empirically quantifiable}.

To validate this conjecture we model 120 different performance traces using a variety of forecasting models which we introduce in Section \ref{sec:model}. We then compare the empirically estimated \emph{complexity} with the accuracy of each forecast method. Which results in two primary findings:

@@

% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU: instructions executed per second, cache misses per 100,000
instructions, and similar statistics. Prediction of these quantities
at some finite time in the future, even with perfect accuracy, does
not violate the Rice-Shapiro theorem.

 The rest of the paper is organized as follows: Section~\ref{related} discusses similar methods and how this method is different. Section~\ref{sec:methods} explains the experimental setup and explains in depth the programs observed to generate each time series. Section~\ref{sec:Model} describes the four forecast models we implemented.  In
 Section~\ref{sec:meaComplex}, we review weighted permutation entropy. In Section~\ref{sec:results} we estimate the complexity of each time series, and
 compare the complexity to the prediction accuracy of several different forecast schema.  In
 Section~\ref{sec:conc}, we discuss these results and their
 implications in regard to our conjecture, and consider future areas of
 research.

>>>>

\section{Introduction {\color{blue} EDITABLE}}\label{sec:intro}

@@

Complicated time series are ubiquitous in modern scientific research. 
Every observed time series, complicated or not, exists on a spectrum of \emph{complexity}. On the low end of that spectrum are time series that exhibit perfect predictive structure, i.e, signals whose future value are perfectly predicted by knowledge of past values from some ``ideal" model. In particular, there is a an underlying process that generates and transmits information from the past to the future in a perfectly predictable fashion. Some examples of this are constant or low-period signals. On the opposite end of this spectrum are signals which are ``fully complex", i.e., the underlying generating process creates information so rapidly that no information at all is transmitted from the past to the future. Some examples, of this are white noise or random walk processes. With signals on this side of the spectrum, knowledge of the past gives absolutely no insight into the future, \emph{regardless} of the chosen model.  In the middle of this spectrum are where complicated \emph{and} interesting---from a modeling perspective---signals exist, e.g., observations of a deterministic chaotic trajectory, high period signals with some level of noise on top, etc. With time series in this portion of the spectrum, enough information is being transmitted from the past to the future that given the \emph{ideal} model the future of the observed system could be forecast with high accuracy.

Unfortunately, as a corollary of the undecidability of the halting problem: there cannot exist \emph{one} universally ideal forecasting scheme for even completely noise-free deterministic time series\cite{weigend-book}, let alone an arbitrary time series. This naturally leads to an important and hard question: given a complicated real-valued time series, does their exist a forecast model for that time series that can utilize the information being transmitted from past to future (if such information exists)? A first step in answering this question is to answer: is it possible to reliably quantify where on the spectrum of complexity the time series lies? i.e, when little or nothing is known about the underlying system's generating process (e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary etc.) can it be inferred without a model if the signal is too \emph{complex} to effectively forecast using a prediction strategy based on predictive structure (temporal forward information transfer)? We rigorously define this \emph{complexity spectrum} as well as what is meant by \emph{low, medium and high} in Section \ref{sec:meaComplex}.

To answer this question and for the purposes of this paper we focus on real-valued (possibly) noisy scalar time series which appear \emph{complicated}, i.e., they exhibit interesting behavior, for instance, not of low period, strictly monotonic, or constant. We make no assumptions about the generating process's properties,(e.g., linear, nonlinear, deterministic, stochastic, stationary, non-stationary, etc.) and we attempt to measure empirically where on the previously mentioned spectrum of \emph{complexity} the time series exists.
We argue that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy of a real-valued-finite-length-noisy time series through ordinal analysis---is an effective way to explore that conjecture. In light of this, for the purposes of this paper we define \emph{complexity} as an approximation of Shannon \emph{entropy}, which we define and explain rigorously in Section \ref{sec:meaComplex}. In this fashion a random walk time series (which exhibits high entropy) is purely complex, where as a periodic signal (which exhibits low entorpy) has low complexity. {\color{blue} Ryan: Insert your sentence here, Something like We are not the first people to attempt to quantify complexity of time series with entropy(citations) However, these methods rely on binning which is bad because of biasing (citations) which we explain further in Section \ref{sec:related}.}

@@

To explore this conjecture we require a test bed for generating a broad array of time series which cover the full spectrum of complexity. We argue that the ideal test bed for this purpose are computer performance traces, e.g., the processor efficiency during the execution of a particular program. Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor efficiency highly complicated. 
%Accurate forecasts of these quantities, if one could
%construct them, could be used to improve computer design. 
% If one
%could predict that a particular computational thread %would be bogged
%down for the next 0.6 seconds waiting for data from main memory, for
%instance, one could save power by putting that thread %on hold for that
%time period (e.g., by migrating it to a processing %unit whose clock
%speed is scaled back).

As a result computer performance traces are very
complicated and range across the entire spectrum of complexity, making this an ideal test bed.  Even a simple ``microkernel," like a three-line loop that
repeatedly initializes the upper triangle of a matrix in column-major order, can produce everything from periodic to
chaotic performance traces \cite{mytkowicz09}, depending on the architecture running the software. Such a performance trace is shown in
Figure~\ref{fig:ipc}.%, and chaos places fundamental limits on predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of \col, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
More interesting programs, e.g., compilers and numerical solvers, produce time series which exist throughout the spectrum of complexity: from completely predictable to completely unstructured,i.e., low to high complexity[[Just trying to reinforce the wording in the readers mind]]. In addition, some performance traces even exhibit interesting regime changes as a program switches between varying subroutines. With a single system (computer performance) capable of producing complicated time series that vary so greatly in complexity (periodic to noise), it would be invaluable to known a priori if a trace contained enough predictive structure to build a  forecast model based on structure before time and effort was spent searching for an ideal model.

@@

This paper provides insight into when, why, and how forecast strategies fail when they are applied to complicated time series by empirically estimating their inherent complexity. [[Liz, do you think here would be better for the sentence about we are not the first, or is above where I asked Ryan to insert it ok? I worry it would muddy the water of this synopsis paragraph.]] We focus here on the specific example of
computer performance but believe the results apply to a much broader class of time series. We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity, as well as from
measurement issues like noise, aggregation, and finite data
length---is \emph{empirically quantifiable}. To validate this conjecture we model 120 different performance traces using a variety of forecasting models which we introduce in Section \ref{sec:model}. We then compare the empirically estimated \emph{complexity} with the accuracy of each forecast method. Which results in two primary findings:

@@

% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU, e.g., processor efficiency and similar statistics. Prediction of these quantities
at some finite time in the future, even with perfect accuracy, does
not violate the Rice-Shapiro{\color{blue}[[Ryan: Do you have a citation for this theorem, also should we give a brief sentence explaining what this theorem is?]]} theorem.

 The rest of the paper is organized as follows: Section~\ref{sec:related} discusses similar methods and how this method is different. Section~\ref{sec:methods} explains the experimental setup and explains in depth the programs observed to generate each time series. Section~\ref{sec:model} describes the four forecast models we implemented.  In
 Section~\ref{sec:meaComplex}, we review weighted permutation entropy. In Section~\ref{sec:results} we estimate the complexity of each time series, and
 compare the complexity to the prediction accuracy of several different forecast schema.  In
 Section~\ref{sec:conc}, we discuss these results and their
 implications in regard to our conjecture, and consider future areas of
 research.

### 2014-03-04 related.tex
edits to intro,related and methods

\section{Related Work}

@@

Entropy though Generating partitions 
works **if** you have generating parition but not if you don't

>>>>

\section{Related Work}\label{sec:related}

@@

Entropy though Generating partitions 
works **if** you have generating parition but not if you don't
{\color{blue} Ryan: Please insert here the paragraph and citations defaming binning / partitioning and illustrating the difficulty of generating partitions / biasing of this method without generating partitions, along with relevant citations. }
